\pdftitle

### 2017-05-18 maxNt.tex
discussion

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-18T14:18:19+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\usepackage[final,nonatbib]{nips_2017}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\pdftitle}{Representative random samples\\ and maximum-entropy
  distributions:\\ a dilemma}
\newcommand*{\headtitle}{\pdftitle}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, V. Rostami, E. Torre}

@@

% LM: I'd like to avoid common beginnings like "Today blah blah is
% important" or "Blah blah is interesting/fundamental". They feel more like
% a mechanical ritual than a real tentative of connecting with the reader.
This note is mainly addressed to neuroscientists interested in \me\
methods, but we would be pleased if its discussion of the probability of
\enquote{random sampling} were useful to neuroscientists that use other
statistical methods to study the physical and dynamical characteristics of
brain areas via neuronal recording.

Recent electrophysiological techniques \cite{berenyietal2014} in fact allow
experimenters to record samples of even a couple hundreds neurons from
specific brain areas. These samples are usually picked out according to an
unknown process, but from their observation we expect to learn something
about all other neurons in the same brain area. That is, we assumed
that they are a \enquote{representative random sample}.

We don't elaborate on the various important purposes of such recordings
here \cite{?***}, but stress that these sample sizes can be considered
\enquote{large}, because their statistical analysis requires considerable
computational power.

These computational costs are one, probably not the earliest, of the many
reasons why \me\ methods have been introduced in neuroscience. It would be
useful to somehow compress the statistical wealth of large neuronal
recordings into few quantities, like sample moments for example. This
compression would also entail interesting physico-biological properties of
neuronal activity. The standard \me\ method
\cite{jaynes1957,sivia1996_r2006,meadetal1984} accomplishes this kind of
compression: it associates a unique probability distribution with few
experimental quantities \cite{***}.
But this is only one of its uses. It is
also used for various information-theoretic purposes or to generate
reference probability distributions \cite{schneidmanetal2006,shlensetal2006,roudietal2009b,tkaciketal2014b}.
In all these uses the \me\
distribution is chosen as the \enquote{maximally noncommittal} one
\cite{jaynes1963}. This adjective means little without further technical
characterizations. Different works give different characterizations, but in
this paper we will use the quoted expression as an umbrella term for all of
them. Our results will not depend on the specific characterization of
\enquote{noncommittal}.

@@

In this note we want to show that we have to face a dilemma if we want to
apply the \me\ method to a representative random sample to find a maximally
noncommittal distribution.

@@

In the rest of the paper we mathematically formulate this dilemma. To this
purpose we also present some probability relations relevant to
\enquote{random sampling}. These relations are well-known in survey
sampling and in the pedagogic problem of drawing from an urn without
replacement, yet they are somewhat hard to find explicitly written in the
neuroscientific literature, so they may be of interest on their own.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=0.99\linewidth]{pop_sample_projection3.pdf}%
\caption{Log-plot of the hypergeometric distribution
  $%\p(\yxs = \yrs \|\yXf=\yRf, \yH)=
\ypp(\yrs\|\yRf) \defd  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}$ for $N=5000$, $n=200$. (Band artifacts may appear in the
  colourbar depending on your \textsc{pdf} viewer.)}
\label{fig:hypergeom_proj}
\end{figure}%hypergeometric_identity.nb
A look at a plot of the hypergeometric distribution $\ypp(\yrs\|\yRf)$, see
\fig~\ref{fig:hypergeom_proj}, reveals that it is a sort of \enquote{fuzzy
  identity matrix} between the $\yXf$-space $\set{0,\dotsc,N}$ and
$\yxs$-space $\set{0,\dotsc,n}$. When $n=N$ it is the identity matrix. We
thus have that
\begin{equation}
  \label{eq:roughly_equal_nN}
  \p(\yxs=a) \approx \p(\yXf=a),\qquad
\expeb{f(\yxs)} \approx \expeb{f(\yXf)}.
\end{equation}
These are only very approximate equalities: they may miss important
features of the two probability distributions. In the next section we will
in fact emphasize their differences. If the distribution for the population
average $\yXf$ is bimodal, for example, the bimodality can be lost in the
distribution for the sample average $\yxs$, owing to the coarsening effect
of $\ypp(\yrs\|\yRf)$.

@@

Suppose we want to constrain the sample expectations of a vector function
$\yf = (f_1,\dotsc,f_m)$ to the vector values $\yc=(c_1,\dotsc,c_m)$, that
is, $\expeb{\yf(\yxs)}=\yc$. Application of \me\
\cite{sivia1996_r2006,meadetal1984} at the population level, denoted by
$\yHa$, gives
\begin{equation}
  \label{eq:app_maxent_pop}
  \p(\yXf = \yRf \| \yHa)  = \yK\,
  \binom{N}{N \yRf}\exp\Bigl[\yL\T
  \sum_{n\yrs=0}^n \yf(\yrs)\ypp(\yrs\|\yRf)\Bigr],
\end{equation}
and then by marginalization with \eqn~\eqref{eq:marginal}
\begin{equation}
  \label{eq:app_maxent_pop_marg}
  \p(\yxs = \yrs \| \yHa)  = \yK\, 
  \sum_{N\yRf=0}^N \ypp(\yrs \|\yRf)
\,
  \binom{N}{N \yRf}\exp\Bigl[\yL\T
  \sum_{n\yrs=0}^n \yf(\yrs)\ypp(\yrs\|\yRf)\Bigr],
\end{equation}
where $\yK$ is a normalization constant and
$\yL\T=(\varLambda_1,\dotsc,\varLambda_m)\T$ are Lagrange multipliers such
that
\begin{equation}
  \label{eq:app_maxent_pop_constraints}
\yc = \yK\sum_{n\yrs=0}^n \sum_{N\yRf=0}^N \yf(\yrs) \ypp(\yrs \|\yRf)
\,
  \binom{N}{N \yRf}\exp\Bigl[\yL\T
  \sum_{n\yrs=0}^n \yf(\yrs)\ypp(\yrs\|\yRf)\Bigr].
\end{equation}
Application of \me\ at the sample level, denoted by $\yHb$, gives
\begin{equation}
  \label{eq:app_maxent_sample}
  \p(\yxs=\yrs \|\yHb)
  =\yk\,
  \binom{n}{n\yrs}
  \exp[\yl\T \yf(\yrs)]
\end{equation}
where $\yk$ is a normalization constant and $\yl\T$ are Lagrange
multipliers such that
\begin{equation}
  \label{eq:app_maxent_sample_constraints}
\yc = \yk\sum_{n\yrs=0}^n  \yf(\yrs) \binom{n}{n\yrs}
  \exp[\yl\T \yf(\yrs)].
\end{equation}

@@

A perfunctory intuitive reasoning seems insufficient for clarifying this
point. Let's express it in the language of the probability calculus.
Suppose we do not know whether the sample is really part of a larger
population: we do not know whether $N=n$ or how large is $N$ otherwise.
Call this state of ignorance $\yHd$. From the point of view of the
probability calculus, this ignorance about $N$ is expressed by assigning a
probability distribution $\p(N \| \yHd)$ that vanishes if $N<n$, since we
know that $N\ge n$. Maintaining our assumption of symmetric ignorance,
probability assignments that do not assume a specific value of $N$ are then
obtained by multiplication of all $N$-dependent probabilities by
$\p(N \| \yHd)$ and subsequent marginalization over $N$ -- technically
speaking, $N$ is a \emph{nuisance parameter} that we eliminate
\cite{jaynes1994_r2003,lindley1965b_r2008,bernardoetal1994_r2000}. The
probability obtained from \me\ at the population level,
\eqn~\eqref{eq:app_maxent_pop_marg} then generalizes to
\begin{equation}
  \label{eq:app_maxent_pop_marg_unknown_N}
  \p(\yxs = \yrs \| \yHd)  = \sum_N \biggl\{\yK_N\, 
  \sum_{N\yRf=0}^N \ypp_N(\yrs \|\yRf)
\,
  \binom{N}{N \yRf}\exp\Bigl[{\yL_N}\T
  \sum_{n\yrs=0}^n \yf(\yrs)\ypp_N(\yrs\|\yRf)\Bigr]\biggr\}
  \,\p(N \| \yHd).
\end{equation}
This is a formidable expression. But the answer to our question, whether
the usual \me\ at the sample level~\eqref{eq:app_maxent_sample} does not
assume anything about a larger population, translates into the precise
mathematical question: are the
distributions~\eqref{eq:app_maxent_pop_marg_unknown_N} and
\eqref{eq:app_maxent_sample} equal, for some choice of $\p(N \| \yHd)$? We
leave this mathematical problem for future work.

@@

In a neuroscientific context we find this assumption very natural, and
therefore the \me\ method at the population level preferable. After all,
also \emph{physical} neuronal-network models usually include some sort of
external input to the neurons, mimicking their embedding in a larger
network.

\iffalse
The quandary we found in the present note has at least two bright sides.
The first is that the two different \me\ applications may give numerically
similar results in some cases. One could thus be used as an approximation
of the other, depending on the circumstances -- \eg\ if probability tails
are unimportant. See \fig~\ref{fig:diff_maxent_pop_sample_realdata} for an
example.
\begin{figure}[!t]
\centering
\includegraphics[width=0.95\columnwidth]{different_maxent_pop_sample_200_realdata_both.pdf}
% 
\caption{Example of discrepancy in $\p(\yxs = \yrs \| \yH)$ constructed by
  \me\ at the population level followed by sample marginalization (blue
  triangles), and at the sample level (red circles). The constraints are
  the same in either case: $\expe{\yxs \|\yH}=0.045$,
  $\expe{\yxxs \|\yH}=0.0025$, and $N=5000$, $n=200$.}
\label{fig:diff_maxent_pop_sample_realdata}
\end{figure}%maxent_pop_or_sample.nb
The second is that \fi The possibility of using two different distributions
is not a physical contradiction. Similar situations arise in statistical
mechanics. It is known that if a system is described by a \me\ Gibbs state,
its subsystems need not be \cite{maesetal1999}. A dilemma quite similar to
ours also appears in the statistical description of the final state of a
non-equilibrium process starting and ending in two equilibrium states: we
can describe our knowledge about the final state either by a Gibbs
distribution, or by the distribution obtained from the Liouville evolution
of the Gibbs distribution assigned to the initial state. The two
descriptions differ -- even though the final \emph{physical} state is
obviously exactly the same \cite[\sect~4]{jaynes1985d_r1993}. The
difference in the two descriptions appears because in one case we can make
sharper predictions about the state thanks to our knowledge of its
preceding dynamics. In this example, though, both distributions are usually
immensely sharp and practically lead to the same predictions. In the cases
considered in this note the difference in predictions may be relevant, and
the ones made by \me\ at the population level are more informed.

\mynote{Importance of the formulae relating sample and population.}

\mynote{Discussion of case when individual $X_i\dotsm X_j$ are constrained.
Likely breakdown of \me\ in this case \cite{portamana2009}. Hints at full
Bayesian treatment of which \me\ is a limit? \cite[***]{mackay1995_r2003}}

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-18T16:09:42+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\usepackage[%final,
nonatbib]{nips_2017}

\newcommand*{\firstdraft}{4 November 2015}

\newcommand*{\pdftitle}{Representative samples\\ and maximum-entropy distributions\\ in neuroscience: a dilemma}

\newcommand*{\headtitle}{\pdftitle} \newcommand*{\pdfauthor}{P.G.L. Porta Mana, V. Rostami, E. Torre}

@@

% LM: I'd like to avoid common beginnings like "Today blah blah is
% important" or "Blah blah is interesting/fundamental". They feel more like
% a mechanical ritual than a real tentative of connecting with the reader.
This note is mainly addressed to neuroscientists interested in \me\
methods, but we would be pleased if its discussion of the probability of
\enquote{sampling} were useful to neuroscientists that use other
statistical methods to study the physical and dynamical characteristics of
brain areas via neuronal recording.

Recent electrophysiological techniques \cite{berenyietal2014} in fact allow
experimenters to record samples of even a couple hundreds neurons from
specific brain areas. These samples are usually picked out according to an
unknown process, but from their observation we expect to learn something
about all other neurons in the same brain area. That is, we assume
that they are a \enquote{representative random sample}.

We don't elaborate on the various important purposes of such recordings
here, but stress that these sample sizes can be considered
\enquote{large}, because their statistical analysis requires considerable
computational power.

These computational costs are one, probably not the earliest, of the many
reasons why \me\ methods have been introduced in neuroscience. It would be
useful to somehow compress the statistical wealth of large neuronal
recordings into few quantities, like sample moments for example. This
compression would also entail interesting physico-biological properties of
neuronal activity. The standard \me\ method
\cite{jaynes1957,sivia1996_r2006,meadetal1984} accomplishes this kind of
compression: it associates a unique probability distribution with few
experimental quantities. But this is only one of its uses. It is also used
for various information-theoretic purposes or to generate reference
probability distributions
\cite{bohteetal2000,schneidmanetal2006,shlensetal2006,roudietal2009b,mackeetal2011,tkaciketal2014b,shimazakietal2015}.
In all these uses the \me\ distribution is chosen as the \enquote{maximally
  noncommittal} one \cite{jaynes1963}. This adjective means little without
further technical characterizations. Different works give different
characterizations, but in this paper we will use the quoted expression as
an umbrella term for all of them. Our results will not depend on the
specific characterization of \enquote{noncommittal}.

@@

In this note we want to show that we have to face a dilemma if we want to
apply the \me\ method to a representative  sample to find a maximally
noncommittal distribution.

@@

In the rest of the paper we mathematically formulate this dilemma. To this
purpose we also present some probability relations relevant to
sampling. These relations are well-known in survey
sampling and in the pedagogic problem of drawing from an urn without
replacement, yet they are somewhat hard to find explicitly written in the
neuroscientific literature, so they may be of interest on their own.

@@

\begin{figure}[!t]
\centering
\includegraphics[width=0.99\linewidth]{pop_sample_projection3.pdf}%
\caption{Log-plot of the hypergeometric distribution
  $%\p(\yxs = \yrs \|\yXf=\yRf, \yH)=
\ypp(\yrs\|\yRf) \defd  \raisebox{0pt}[0pt][1ex]{}\binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1}$ for $N=5000$, $n=200$. (Band artifacts may appear in the
  colourbar depending on your \textsc{pdf} viewer.)}
\label{fig:hypergeom_proj}
\end{figure}%hypergeometric_identity.nb
A look at a plot of the hypergeometric distribution $\ypp(\yrs\|\yRf)$, see
\fig~\ref{fig:hypergeom_proj}, reveals that it is a sort of \enquote{fuzzy
  identity matrix} between the $\yXf$-space $\set{0,\dotsc,N}$ and
$\yxs$-space $\set{0,\dotsc,n}$. When $n=N$ it is the identity matrix. We
thus have that
\begin{equation}
  \label{eq:roughly_equal_nN}
  \p(\yxs=a) \approx \p(\yXf=a),\qquad
\expeb{f(\yxs)} \approx \expeb{f(\yXf)}.
\end{equation}
These are only very approximate equalities: they may miss important
features of the two probability distributions. In the next section we will
in fact emphasize their differences. If the distribution for the population
average $\yXf$ is bimodal, for example, the bimodality can be lost in the
distribution for the sample average $\yxs$, owing to the coarsening effect
of $\ypp(\yrs\|\yRf)$.

@@

Suppose we want to constrain the sample expectations of a vector function
$\yf = (f_1,\dotsc,f_m)$ to the vector values $\yc=(c_1,\dotsc,c_m)$, that
is, $\expeb{\yf(\yxs)}=\yc$. Application of \me\
\cite{sivia1996_r2006,meadetal1984} at the population level, denoted by
$\yHa$, gives
\begin{equation}
  \label{eq:app_maxent_pop}
  \p(\yXf = \yRf \| \yHa)  = \yK\,
  \binom{N}{N \yRf}\,\exp\Bigl[\yL\T
  \sum_{n\yrs=0}^n \yf(\yrs)\ypp(\yrs\|\yRf)\Bigr],
\end{equation}
and then by marginalization with \eqn~\eqref{eq:marginal}
\begin{equation}
  \label{eq:app_maxent_pop_marg}
  \p(\yxs = \yrs \| \yHa)  = \yK\, 
  \sum_{N\yRf=0}^N \ypp(\yrs \|\yRf)
\,
  \binom{N}{N \yRf}\,\exp\Bigl[\yL\T
  \sum_{n\yrs=0}^n \yf(\yrs)\ypp(\yrs\|\yRf)\Bigr],
\end{equation}
where $\yK$ is a normalization constant and
$\yL\T=(\varLambda_1,\dotsc,\varLambda_m)\T$ are Lagrange multipliers such
that
\begin{equation}
  \label{eq:app_maxent_pop_constraints}
\yc = \yK\sum_{n\yrs=0}^n \sum_{N\yRf=0}^N \yf(\yrs) \ypp(\yrs \|\yRf)
\,
  \binom{N}{N \yRf}\,\exp\Bigl[\yL\T
  \sum_{n\yrs=0}^n \yf(\yrs)\ypp(\yrs\|\yRf)\Bigr].
\end{equation}
Application of \me\ at the sample level, denoted by $\yHb$, gives
\begin{equation}
  \label{eq:app_maxent_sample}
  \p(\yxs=\yrs \|\yHb)
  =\yk\,
  \binom{n}{n\yrs}
  \exp[\yl\T \yf(\yrs)]
\end{equation}
where $\yk$ is a normalization constant and $\yl\T$ are Lagrange
multipliers such that
\begin{equation}
  \label{eq:app_maxent_sample_constraints}
\yc = \yk\sum_{n\yrs=0}^n  \yf(\yrs) \binom{n}{n\yrs}
  \exp[\yl\T \yf(\yrs)].
\end{equation}

@@

A perfunctory intuitive reasoning seems insufficient for clarifying this
point. Let's express it in the language of the probability calculus.
Suppose we do not know whether the sample is really part of a larger
population: we do not know whether $N=n$ or how large is $N$ otherwise.
Call this state of ignorance $\yHd$. In the probability calculus, this
ignorance about $N$ is expressed by assigning a probability distribution
$\p(N \| \yHd)$ that vanishes if $N<n$, since we know that $N\ge n$; see
Good \cite{good1965,good1967b} and Rissanen \cite{rissanen1983} for
examples of such distributions over the integers. Maintaining our
assumption of symmetric ignorance, probability assignments that do not
assume a specific value of $N$ are then obtained by multiplication of all
$N$-dependent probabilities by $\p(N \| \yHd)$ and subsequent
marginalization over $N$. Technically speaking, $N$ becomes a
\emph{nuisance parameter}
\cite{jaynes1994_r2003,lindley1965b_r2008,bernardoetal1994_r2000}. The
probability obtained from \me\ at the population level,
\eqn~\eqref{eq:app_maxent_pop_marg}, then generalizes to
\begin{equation}
  \label{eq:app_maxent_pop_marg_unknown_N}
  \p(\yxs = \yrs \| \yHd)  = \sum_N \biggl\{\yK_N\, 
  \sum_{N\yRf=0}^N \ypp_N(\yrs \|\yRf)
\,
  \binom{N}{N \yRf}\,\exp\Bigl[{\yL_N}\T
  \sum_{n\yrs=0}^n \yf(\yrs)\ypp_N(\yrs\|\yRf)\Bigr]\biggr\}
  \,\p(N \| \yHd),
\end{equation}
where $N$-dependencies have been made explicit. This is a formidable
expression. But the answer to our question, whether the usual \me\ at the
sample level~\eqref{eq:app_maxent_sample} does not assume anything about a
larger population, translates now into the precise mathematical question:
are the distributions~\eqref{eq:app_maxent_pop_marg_unknown_N} and
\eqref{eq:app_maxent_sample} equal for some choice of $\p(N \| \yHd)$, with
$\p(N \| \yHd)\ne 0$ for $N>n$? We leave this mathematical problem for
future work. Note, however, that this equality is satisfied if
$\p(N=1 \| \yHd)=1$, which means that the usual \me\ model can also be
interpreted as assuming that \emph{no} larger population exists.

@@

Rough estimates of $N$ may be available in neuroscientific contexts, so we
find the \me\ model constructed at the population level very natural and
preferable. After all, \emph{physical} neuronal-network models usually
include some sort of external input to the neurons as well, mimicking their
embedding in a larger network. The distribution of this \me\ model, when
used as a reference distribution for surprise analysis, may reveal features
in a dataset that were unnoticed by the standard \me\ model.

\iffalse
The quandary we found in the present note has at least two bright sides.
The first is that the two different \me\ applications may give numerically
similar results in some cases. One could thus be used as an approximation
of the other, depending on the circumstances -- \eg\ if probability tails
are unimportant. See \fig~\ref{fig:diff_maxent_pop_sample_realdata} for an
example.
\begin{figure}[!t]
\centering
\includegraphics[width=0.95\columnwidth]{different_maxent_pop_sample_200_realdata_both.pdf}
% 
\caption{Example of discrepancy in $\p(\yxs = \yrs \| \yH)$ constructed by
  \me\ at the population level followed by sample marginalization (blue
  triangles), and at the sample level (red circles). The constraints are
  the same in either case: $\expe{\yxs \|\yH}=0.045$,
  $\expe{\yxxs \|\yH}=0.0025$, and $N=5000$, $n=200$.}
\label{fig:diff_maxent_pop_sample_realdata}
\end{figure}%maxent_pop_or_sample.nb
The second is that \fi

The possibility of using two different distributions is not a physical
contradiction. Similar situations arise in statistical mechanics. It is
known that if a system is described by a \me\ Gibbs state, its subsystems
need not be \cite{maesetal1999}. A dilemma quite similar to ours also
appears in the statistical description of the final state of a
non-equilibrium process starting and ending in two equilibrium states: we
can describe our knowledge about the final state either by a Gibbs
distribution, or by the distribution obtained from the Liouville evolution
of the Gibbs distribution assigned to the initial state. The two
descriptions differ -- even though the final \emph{physical} state is
obviously exactly the same \cite[\sect~4]{jaynes1985d_r1993}. The
difference in the two descriptions appears because in one case we can make
sharper predictions about the state thanks to our knowledge of its
preceding dynamics. In this example, though, both distributions are usually
immensely sharp and practically lead to the same predictions. In the cases
considered in this note the difference in predictions may be relevant
instead.

Our analysis touched only constraints of the sample average. The
corresponding models are usually called \enquote{homogeneous} in the
literature. Purely \enquote{inhomogeneous} models have also been used
\cite{schneidmanetal2006,shlensetal2006,roudietal2009b}, in which
expectations for individual neurons or groups of neurons are constrained,
for example $\expe{x_2}$ or $\expe{x_1x_8x_9}$. A short computation shows
that the \me\ method with this kind of constraints gives the same result
whether applied at the sample or at the population level: the states of any
unconstrained neurons marginalize out. This is understandable: expressing
different uncertainties about neurons $2$ and $5$ we are breaking the
symmetry of our uncertainty, which thus cannot be representative of other
neurons in the sample or in the population.

Inhomogeneous models, however, require enormous computational power for
large sample sizes; homogeneous models therefore retain their importance.
Our analysis and dilemma also persist for hybrid homogeneous-inhomogeneous
models \cite{tkaciketal2014b,shimazakietal2015}.

%\mynote{Importance of the formulae relating sample and population.}

% \mynote{Discussion of case when individual $X_i\dotsm X_j$ are constrained.
% Likely breakdown of \me\ in this case \cite{portamana2009}. Hints at full
% Bayesian treatment of which \me\ is a limit? \cite[***]{mackay1995_r2003}}
